---
title: Using Data from different Packages
slug: dataset-in-r
author: Junali Chhetri
date: 2017-03-01
categories:
  - R
---
Most of the time we need data either for teaching purpose or learning things. Whatever be the reason, data are really important things. R and different packages of R contains different datasets which we can use directly just by installing them. Many books related to R illustrate there examples using these data sources. We can find the dataset from differnt sources as `data()`

The datasets from the package `datasets` and other loaded packages is shown in default. However, we can obtain the datasets contained in specific packages as,

```{r eval = FALSE}
data(package='MASS')
```

The above will output the dataset contained in the package `MASS`. These datasets can be saved in a variable and used as different data structure like matrix, data frame or may be list. For example,

```{r eval = FALSE}
data(Rubber, package='MASS') 
rubber.data <- Rubber
```


### Exercises
  1. Save the Rabbit dataset from the package `MASS` into rabbitData and find the mean of BPchange for different Doses applied.
  2. Load the data Greene from package `car` and find which location has the maximum variation in success.
