---
title  : How ANOVA analyze the variance
slug   : how-anova-analyze-the-variance
author : therimalaya
date   : 2021-03-29
tags:
  - Statistics
  - ANOVA
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  comment = NA,
  out.width = "100%",
  dev = "svg"
)
library(data.table)
library(purrr)
library(ggplot2)

ggplot2::theme_set(ggthemes::theme_few() + theme(
  panel.grid = element_line(color = "#f0f0f0")
))

set.seed(124)
```


Often we Analysis of Variance (ANOVA) to analyze the variances to find if 
different cases results in similar outcome and if the difference is significant. 
Following are some simple examples,

- The effect of different diets on growth of fishes
- Comparing the height of three different species of a plant
- Type of flour used for baking a bread

These are some common examples where in some cases data are collected by setting 
up an experiment and in other cases they are collected through sampling. 
This article tries to explain how the ANOVA analyze the variance and in what 
situation are they significant throught both simulated and real data.

Consider the following model with $i=3$ groups and $j=n$ observations,

$$y_{ij} = \mu + \tau_i + \varepsilon_{ij}, \; i = 1, 2, 3 
\texttt{ and } j = 1, 2, \ldots n$$

here, $\tau_i$ is the effect corresponding to group $i$ and 
$\varepsilon_{ij} \sim \mathrm{N}(0, \sigma^2)$, the usual assumption of 
linear model. In order to understand how ANOVA finds the differences between 
groups and how the group mean and their standard deviation influence the
results from ANOVA, let us consider following four cases:

Case 1
: Similar group means with high variation within the groups

Case 2
: Similar group means with low variation within the groups

Case 3
: Distant group means with high variation within the groups

Case 4
: Distant group means with low variation within the groups

## Simulating data resembling these cases

```{r}
Design <- rbindlist(list(
    Case1 = data.table(
        Group = paste("G", 1:2, sep = ""),
        Mean = c(6, 8),
        SD = c(6, 8)
    ),
    Case2 = data.table(
        Group = paste("G", 1:2, sep = ""),
        Mean = c(6, 8),
        SD = c(3, 4)
    ),
    Case3 = data.table(
        Group = paste("G", 1:2, sep = ""),
        Mean = c(8, 20),
        SD = c(8, 20)
    ),
    Case4 = data.table(
        Group = paste("G", 1:2, sep = ""),
        Mean = c(8, 20),
        SD = c(4, 10)
    )
), idcol = "Cases")
```

## Fitting ANOVA model for each cases
```{r}
Model <- Design[, map2_df(Mean, SD, ~list(data.table(
    Response = rnorm(25, .x, .y)
))), by = .(Cases, Group)] %>% 
.[, .(Data = list(.SD)), by = Cases] %>% 
.[, Fit := map(Data, ~lm(Response ~ Group, data = .x))]
```

## Prepare data for plotting

```{r}
plot_data <- Model[, map_df(Fit, broom::augment), by = Cases]
```


```{r}
ggplot(plot_data, aes(Response, Group)) +
    geom_point(aes(fill = Group),
        position = position_jitter(height = 0.1),
        shape = 21, size = 2, stroke = 0.25, alpha = 0.25
    ) +
    stat_summary(
        fun.data = mean_se, geom = "errorbar", aes(color = Group),
        width = 0.2, size = 0.75
    ) +
    stat_summary(
        fun = mean, geom = "point", aes(color = Group),
        size = 2, shape = 21, fill = "whitesmoke", stroke = 0.75
    ) +
    facet_wrap(facets = vars(Cases), scales = "free_x") +
    ggridges::geom_density_ridges(
        aes(color = Group),
        fill = NA,
        panel_scaling = FALSE
    ) +
    scale_color_brewer(palette = "Set1") +
    scale_fill_brewer(palette = "Set1")
```

```{r}
ggplot(plot_data) +
    geom_text(
        data = Design, aes(
            label = glue::glue("Mean: {Mean}, SD: {SD}"),
            color = Group
        ),
        x = Inf, y = Inf,
        hjust = 1.1,
        vjust = rep(c(1.5, 3.5), 4),
        show.legend = FALSE
    ) +
    geom_point(
        aes(Response, Group, color = Group),
        position = position_jitter(height = 0.1),
        shape = 21, stroke = 0.5,
        alpha = 0.5
    ) +
    ggridges::geom_density_ridges(
        aes(Response, Group, color = Group),
        scale = 0.9,
        fill = NA
    ) +
    ggridges::geom_density_ridges(
        aes(.fitted, Group, color = Group), 
        linetype = "dashed",
        scale = 0.9,
        fill = NA
    ) +
    facet_wrap(
        facets = vars(Cases),
        dir = "v",
        scales = "free_x"
    ) +
    scale_color_brewer(palette = "Set1") +
    theme(
        legend.position = "bottom"
    ) +
    labs(
        x = "Response",
        y = "Density",
        linetype = NULL
    )
```

## Discussion



