---
title: "Simulating data for ANOVA similar to existing dataset for analysis"
author: Raju Rimal
date: "`r Sys.Date()`"
draft: true
tags:
  - Statistics
  - ANOVA
---

Simulating data is used in education and research and has been an helpful tool
for testing, comparing and understanding concepts in practical and applied 
setting.


```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  comment = NA,
  out.width = "100%",
  dev = "svg"
)
library(data.table)
library(purrr)
library(ggplot2)

ggplot2::theme_set(ggthemes::theme_few() + theme(
  panel.grid = element_line(color = "#f0f0f0")
))

set.seed(124)
```


Often we Analysis of Variance (ANOVA) to analyze the variances to find if 
different cases results in similar outcome and if the difference is significant. 
Following are some simple examples,

- The effect of different diets on growth of fishes
- Comparing the height of three different species of a plant
- Type of flour used for baking a bread

These are some common examples where in some cases data are collected by setting 
up an experiment and in other cases they are collected through sampling. 
This article tries to explain how the ANOVA analyze the variance and in what 
situation are they significant throught both simulated and real data.

Consider the following model with $i=3$ groups and $j=n$ observations,

$$y_{ij} = \mu + \tau_i + \varepsilon_{ij}, \; i = 1, 2, 3 
\texttt{ and } j = 1, 2, \ldots n$$

here, $\tau_i$ is the effect corresponding to group $i$ and 
$\varepsilon_{ij} \sim \mathrm{N}(0, \sigma^2)$, the usual assumption of 
linear model. Simulation example below describe it in details.

# Simulation Example
This example simulates 3 datasets within each of them has 3 groups with
with their respective mean and standard deviation. These datasets tries to 
replicate the `USArrest` dataset available in `dataset` package in R. The 
simulation design is as follows,

```{r, echo = TRUE, class.source = c("bg-info", "folded")}
sim_design <- data.table(
  Illitracy = factor(
    rep(c(1, 2, 3), 3),
    labels = c("high", "medium", "low")
  ),
  Crime = factor(
    rep(c(1, 2, 3), each = 3),
    labels = c("Murder", "Assault", "Rape")
  ),
  mean = c(11, 8, 5, 214, 190, 114, 23, 21, 19),
  sd = c(3, 4, 3, 79, 82, 55, 8, 10, 10)
)

sim_design
```

Since these data cannot contain negative values so instead of using `rnorm` 
available in `stats` package, I will use `truncnorm` available in GitHub. There 
are other options as well which can be used such as: ...

If not installed, install the package as 
`remotes::install_github("olafmersmann/truncnorm")` or 
`devtools::install_github("olafmersmann/truncnorm")`.

In each group let's simulate 50 observations.

```{r, echo = TRUE}
nsim <- 50
```

```{r, echo = TRUE, class.source = c("bg-info", "folded")}
sim_data <- copy(sim_design)[, map2_df(mean, sd, ~ data.table(
  Rate = truncnorm::rtruncnorm(
    n = nsim, a = 0, b = Inf, mean = .x, sd = .y
  ) %>% round()
)), by = .(Illitracy, Crime)]

sim_data
```

Lets use this simulated dataset for fitting Anova model with `Illitracy`
as the factor (group) variable and `Rate` as the response variable 
separately for each crime. Also make a density plot for both simulated and 
fitted values of the `Rate` variable.

## R-codes {.tabset}

### Model and plot data {.tab}
```{r, echo = TRUE, class.source = c("bg-info", "folded")}
mdl_fit <- sim_data[, .(
  Fit = list(lm(Rate ~ Illitracy, data = .SD))
), by = .(Crime)][, Summary := map(Fit, summary)]
mdl_fit[, Anova := map(Fit, anova)]
mdl_fit[, Tukey := map(Fit, aov) %>% map(TukeyHSD)]

mdl_est <- mdl_fit[, map_df(
  .x = Summary,
  .f = broom::tidy,
  conf.int = TRUE
), by = .(Crime)]

mdl_fit_df <- mdl_fit[, map_df(
  .x = Fit,
  .f = broom::augment
), by = .(Crime)]

eff_df <- mdl_fit[, map_df(Fit, function(.fit) {
  effects::Effect("Illitracy", .fit) %>%
    as.data.table()
}), by = "Crime"]

tky_df <- mdl_fit[, modify_depth(Tukey, 1, "Illitracy") %>%
  as.data.table(keep.rownames = "terms"), by = "Crime"]
```

### Density plot {.tab}
```{r, echo = TRUE, class.source = c("bg-info", "folded")}
density_plts <- split(sim_data, by = "Crime") %>%
  imap(function(dta, name) {
    ggplot(dta, aes(Rate, color = Illitracy)) +
      geom_density(aes(linetype = "Simulated")) +
      geom_density(
        aes(linetype = "Fitted", x = .fitted),
        data = mdl_fit_df[Crime == name]
      ) +
      geom_rug(
        data = eff_df[Crime == name],
        aes(x = fit)
      ) +
      scale_linetype_manual(
        breaks = c("Simulated", "Fitted"),
        values = c("solid", "dashed")
      ) +
      scale_color_brewer(palette = "Set1") +
      theme(legend.position = "bottom") +
      labs(
        x = name,
        y = "Density",
        linetype = NULL
      )
  })
```

### Box plot {.tab}
```{r, echo = TRUE, class.source = c("bg-info", "folded")}
box_plts <- split(mdl_fit_df, by = "Crime") %>%
  imap(function(dta, name) {
    ggplot(dta, aes(Rate, Illitracy)) +
      geom_boxplot(
        notch = TRUE, 
        color = "grey",
        outlier.colour = "grey"
      ) +
      geom_point(
        position = position_jitter(height = 0.25),
        color = "grey",
      ) +
      geom_pointrange(
        aes(
          color = "Estimated",
          xmin = lower,
          xmax = upper,
          x = fit
        ),
        data = eff_df[Crime == name]
      ) +
      geom_point(
        aes(color = "True", x = mean),
        data = sim_design[Crime == name]
      ) +
      scale_color_brewer(
        name = "Mean",
        palette = "Set1"
      ) +
      theme(
        legend.position = "bottom"
      )
  })
```

### Posthoc plot {.tab}
```{r, echo = TRUE, class.source = c("bg-info", "folded")}
tky_plts <- split(tky_df, by = "Crime") %>%
  imap(function(dta, name) {
    ggplot(dta, aes(diff, terms)) +
      geom_pointrange(
        aes(xmin = lwr, xmax = upr, x = diff),
        shape = 21,
        fill = "whitesmoke"
      ) +
      geom_vline(
        xintercept = 0,
        linetype = "dashed",
        color = "royalblue"
      ) +
      scale_color_brewer(
        name = "Mean",
        palette = "Set1"
      ) +
      labs(
        y = "Illitracy",
        x = "Effect difference",
        title = "Pairwise comparison of levels of illitracy"
      ) +
      expand_limits(x = 0)
  })
```

## Analysis {.tabset}

### Murder {.tab .tabset}

#### Distribution {.tab}

```{r}
density_plts[["Murder"]]
```

#### Fit {.tab}

```{r}
mdl_fit[Crime == "Murder", Summary][[1]]
```

#### Effects {.tab}

```{r}
box_plts[["Murder"]]
```

#### Post-hoc {.tab}
```{r}
tky_plts[["Murder"]]
```

### Assault {.tab .tabset}

#### Distribution {.tab}

```{r, out.width="100%"}
density_plts[["Assault"]]
```

The distribution of the high and medium illitracy is similar as contrast to low 
illitracy group in the case of assult.

#### Fit {.tab}

```{r}
mdl_fit[Crime == "Assault", Summary][[1]]
```

#### Effects {.tab}

```{r}
box_plts[["Assault"]]
```

#### Post-hoc {.tab}
```{r}
tky_plts[["Assault"]]
```


### Rape {.tab .tabset}

#### Distribution {.tab}

```{r}
density_plts[["Rape"]]
```

#### Fit {.tab}

```{r}
mdl_fit[Crime == "Rape", Summary][[1]]
```

#### Effects {.tab}

```{r}
box_plts[["Rape"]]
```

#### Post-hoc {.tab}
```{r}
tky_plts[["Rape"]]
```

# Real Data Example

```{r, echo = TRUE, class.source = c("bg-info", "folded")}
arrest <- merge.data.table(
  as.data.table(USArrests, keep.rownames = "States"),
  as.data.table(
    state.x77[, "Illiteracy", drop = FALSE],
    keep.rownames = "States"
  ),
  by = "States"
)
arrest[, UrbanPop := NULL]
arrest <- arrest[, lapply(.SD, as.numeric), by = "States"]

arrest[, Illiteracy := cut.default(
  Illiteracy,
  breaks = quantile(Illiteracy, c(0, 1/3, 2/3, 1)),
  labels = c("low", "medium", "high"),
  include.lowest = TRUE
)]

arrest <- melt.data.table(
  data = arrest[, lapply(.SD, as.numeric), 
    by = .(States, Illiteracy)],
  id.vars = c("States", "Illiteracy"),
  value.name = "Rate",
  variable.name = "Crime"
)
```

```{r}
head(arrest)
```

## Results {.tabset}

```{r}
mdl_fit <- arrest[, .(
  Fit = list(lm(Rate ~ Illiteracy, data = .SD))
), by = .(Crime)][, Summary := map(Fit, summary)]
mdl_fit[, Anova := map(Fit, anova)]
mdl_fit[, Tukey := map(Fit, aov) %>% map(TukeyHSD)]

mdl_est <- mdl_fit[, map_df(
  .x = Summary,
  .f = broom::tidy,
  conf.int = TRUE
), by = .(Crime)]

mdl_fit_df <- mdl_fit[, map_df(
  .x = Fit,
  .f = broom::augment
), by = .(Crime)]

eff_df <- mdl_fit[, map_df(Fit, function(.fit) {
  effects::Effect("Illiteracy", .fit) %>%
    as.data.table()
}), by = "Crime"]

tky_df <- mdl_fit[, modify_depth(Tukey, 1, "Illiteracy") %>%
  as.data.table(keep.rownames = "terms"), by = "Crime"]
```

#### Distribution {.tab}

```{r, out.width="100%", fig.asp=0.5}
ggplot(arrest, aes(Rate, color = Illiteracy)) +
  geom_density(aes(linetype = "Simulated")) +
  geom_density(
    aes(linetype = "Fitted", x = .fitted),
    data = mdl_fit_df
  ) +
  geom_rug(
    data = eff_df,
    aes(x = fit)
  ) +
  scale_linetype_manual(
    breaks = c("Simulated", "Fitted"),
    values = c("solid", "dashed")
  ) +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  labs(
    x = "Crime",
    y = "Density",
    linetype = NULL
  ) +
  facet_wrap(
    facets = vars(Crime),
    scales = "free"
  )
```

#### Fit {.tab .tabset}

##### Murder {.tab}
```{r}
mdl_fit[Crime == "Murder", Summary][[1]]
```

##### Assult {.tab}
```{r}
mdl_fit[Crime == "Assault", Summary][[1]]
```

##### Rape {.tab}
```{r}
mdl_fit[Crime == "Rape", Summary][[1]]
```

#### Effects {.tab}

```{r, out.width="100%", fig.asp=0.5}
ggplot(arrest, aes(Rate, Illiteracy)) +
  geom_boxplot(
    notch = FALSE,
    color = "grey",
    outlier.colour = "grey"
  ) +
  geom_point(
    position = position_jitter(height = 0.25),
    color = "grey",
  ) +
  geom_pointrange(
    aes(
      xmin = lower,
      xmax = upper,
      x = fit
    ),
    color = "firebrick",
    data = eff_df
  ) +
  scale_color_brewer(
    name = "Mean",
    palette = "Set1"
  ) +
  theme(
    legend.position = "bottom"
  ) +
  facet_wrap(facets = vars(Crime), scales = "free_x")
```

#### Post-hoc {.tab}
```{r, out.width="100%", fig.asp=0.5}
ggplot(tky_df, aes(diff, terms)) +
  geom_pointrange(
    aes(xmin = lwr, xmax = upr, x = diff),
    shape = 21,
    fill = "whitesmoke"
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "royalblue"
  ) +
  scale_color_brewer(
    name = "Mean",
    palette = "Set1"
  ) +
  labs(
    y = "Illiteracy",
    x = "Effect difference",
    title = "Pairwise comparison of levels of illiteracy"
  ) +
  expand_limits(x = 0) +
  facet_wrap(facets = vars(Crime), scales = "free_x")
```

## Discussion



